---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
Unzip and prepare data
```{r data, cache=TRUE, message=FALSE}
library(dplyr)
unzip("activity.zip")
activity <- tbl_df(read.csv("activity.csv", stringsAsFactors = FALSE))
activity$date <- as.Date(activity$date, "%Y-%m-%d")
str(activity)
summary(activity)
```

## What is mean total number of steps taken per day?
### 1. Calculate the total number of steps taken per day
For this part of the assignment, you can ignore the missing values in the dataset
```{r mean_and_median, cache=TRUE}
per_day <- activity %>%
    group_by(date) %>%
    summarise(tot_steps = sum(steps, na.rm = T))
```

### 2. Make a histogram of the total number of steps taken each day

```{r histogrma_total_steps_per_day_with_mean_and_median, fig.width=12, message=FALSE}
require(dplyr)
steps <- per_day %>% filter(tot_steps > 0)
hist(per_day$tot_steps, main = "Histogram of the total number of steps taken each day", xlab = "Total steps")
abline(v = mean(per_day$tot_steps), col = "red", lwd = 3)
legend("topright", legend = c("mean"), col = c("red"), pch = "-", lwd = 3)
```

### 3. Calculate and report the mean and median of the total number of steps taken per day
Draw total number of steps per day with mean and median
```{r barplot_total_steps_per_day_with_mean_and_median, fig.width=12}
barplot(per_day$tot_steps, names.arg = per_day$date, cex.names = 0.9, las = 2, main = "Mean and median number of steps taken each day")
abline(h=mean(per_day$tot_steps, na.rm = T), col="red", lwd = 3)
abline(h=median(per_day$tot_steps, na.rm = T), col="blue", lwd = 3)
legend("topright", legend = c("mean", "median"), col = c("red", "blue"), pch = "-", lwd = 3)
```

## What is the average daily activity pattern?
### 1. Make a time series plot (i.e. \color{red}{\verb|type = "l"|}type="l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)
```{r prepare_time_intervals, message=FALSE, fig.width=12}
require(dplyr)
per_interval <- activity %>% group_by(interval) %>% summarise(avg_steps = mean(steps, na.rm = TRUE))
plot(per_interval$interval, per_interval$avg_steps, type = "l", 
     main = "Average daily activity ", xlab = "5 min intervals", ylab = "Average steps accross all days")
```

### 2. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
```{r max_steps}
max_avg <- filter(per_interval, avg_steps == max(avg_steps))
```

Maximum average number of steps **`r round(max_avg$avg_steps, 3)`** were done in interval **#`r max_avg$interval`**

## Imputing missing values
Note that there are a number of days/intervals where there are missing values (coded as NA). The presence of missing days may introduce bias into some calculations or summaries of the data.

### 1. Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)
```{r missing_data}
na_count <- sum(is.na(activity$steps))
```
Number if records with missing step values is **`r na_count`**

### 2. Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.
We are going to re-use per_interval data frame to derive avarage steps for missing value per 5-minute interval.
```{r na_replacement, cache=TRUE, message=FALSE}
adjusted_activity <- left_join(activity, per_interval) %>% mutate(new_val = ifelse(is.na(steps), avg_steps, steps))
adjusted_per_day <- adjusted_activity %>%
    group_by(date) %>%
    summarise(tot_steps = sum(new_val))
```
### 3. Create a new dataset that is equal to the original dataset but with the missing data filled in.
```{r new_activity_summary}
str(adjusted_activity)
```

### 4. Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day.
```{r plot_new_hist_and_mean_median_per_day, fig.width=12}
hist(adjusted_per_day$tot_steps, main = "Adjusted histogram of the total number of steps taken each day", xlab = "Total steps")
abline(v = mean(adjusted_per_day$tot_steps), col = "red", lwd = 3)
legend("topright", legend = c("mean"), col = c("red"), pch = "-", lwd = 3)

summary(adjusted_per_day$tot_steps)
barplot(adjusted_per_day$tot_steps, names.arg = adjusted_per_day$date, cex.names = 0.9, las = 2, main = "Adjusted mean and median number of steps taken each day")
abline(h=mean(adjusted_per_day$tot_steps, na.rm = T), col="red", lwd = 3)
abline(h=median(adjusted_per_day$tot_steps, na.rm = T), col="blue", lwd = 3)
legend("topright", legend = c("mean", "median"), col = c("red", "blue"), pch = "-", lwd = 3)
```

In comapison with original dataset the adjusted data looks more normal distributed.

## Are there differences in activity patterns between weekdays and weekends?
For this part the weekdays() function may be of some help here. Use the dataset with the filled-in missing values for this part.

### 1. Create a new factor variable in the dataset with two levels – “weekday” and “weekend” indicating whether a given date is a weekday or weekend day.
```{r weekday_df, cache=FALSE}
weekday_steps <- adjusted_activity %>%
    mutate(wday = ifelse(weekdays(date) %in% c("Saturday", "Sunday"), "weekend", "workday")) %>%
    group_by(interval, wday) %>%
    summarise(avg_steps = mean(new_val, na.rm = T))
```

### 2. Make a panel plot containing a time series plot type="l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis). 
```{r panel_plot, fig.width=12}
library(ggplot2)
qplot(interval, avg_steps, data = weekday_steps, geom = "line", facets=wday~.)
```